A Fully Abstract Game Semantics for Countable Nondeterminism

### 2018-04-16 cones/paper/lipics-v2018-authors/cones.tex
Got it under 15 pages.

\newcommand{\nat}{{\mathtt{nat}}}
\newcommand*\from{\colon}
\newcommand{\Y}{\mathbf{Y}}
\newcommand{\opto}{\longrightarrow}
\newcommand{\n}{{\mathtt{n}}}
\newcommand{\IfO}{{\mathtt{If0}\;}}
\newcommand{\suc}{{\mathtt{suc\;}}}
\newcommand{\pred}{{\mathtt{pred\;}}}
\newcommand{\0}{{\mathtt{0}}}
\newcommand{\com}{{\mathtt{com}}}
\newcommand{\skipp}{{\mathsf{skip}}}
\newcommand{\neww}{{\mathsf{new}}}
\newcommand{\mkvar}{{\mathsf{mkvar}\;}}
\newcommand{\deref}{\texttt{@}}
\newcommand{\ia}[2]{\langle #1 , #2 \rangle}
\newcommand{\stup}[3]{\langle #1 \mid #2 \mapsto #3 \rangle}
\newcommand{\Var}{{\mathtt{Var}}}
\newcommand{\new}{{\mathtt{new}}}
\newcommand{\case}{{\mathtt{case}}}
\newcommand{\blank}{\;\underline{\hspace{1.5ex}}\;}
\newcommand{\bN}{\mathbb{N}}
\newcommand{\bC}{\mathbb{C}}
\newcommand{\deno}[1]{\left\llbracket#1\right\rrbracket}
\newcommand{\converges}{\Downarrow}
\newcommand{\mustconverge}{\converges^{\text{must}}}
\newcommand{\Iflt}{\mathtt{If{<}\;}}
\newcommand{\Ifgt}{\mathtt{If{>}\;}}
\newcommand{\inr}{{\mathsf{inr}}}
\newcommand{\inl}{{\mathsf{inl}}}
\newcommand{\lmam}{\sqsubseteq_{m\&m}}
\newcommand{\emam}{\equiv_{m\&m}}
\newcommand{\lst}{\lesssim}
\newcommand{\rulename}[1]{\LeftTirNameStyle{#1}}
\newcommand{\ts}{{\;\vdash}}
\newcommand{\nts}{{\;\not\vdash}}
\newcommand{\OP}{\{O,P\}}
\newcommand{\QA}{\{Q,A\}}
\DeclareMathOperator{\id}{id}
\DeclareMathOperator{\pr}{pr}
\newcommand{\suchthat}{\;\colon\;}
\newcommand{\varsuchthat}{\;\mid\;}
\newcommand{\esuchthat}{\;.\;}
\newcommand{\emptyplay}{\epsilon}
\newcommand{\tensor}{\otimes}
\renewcommand{\implies}{\multimap}
\newcommand{\prefix}{\sqsubseteq}
\newcommand{\s}{\mathfrak s}
\newcommand{\C}{{\mathcal{C}}}
\newcommand{\G}{\mathcal G}
\DeclareMathOperator{\Int}{int}
\newcommand{\dv}{{\lightning}}
\DeclareMathOperator{\pocl}{pocl}
\newbox\gnBoxA
\newdimen\gnCornerHgt
\setbox\gnBoxA=\hbox{$\ulcorner$}
\global\gnCornerHgt=\ht\gnBoxA
\newdimen\gnArgHgt
\def\pv #1{%
    \setbox\gnBoxA=\hbox{$#1$}%
    \gnArgHgt=\ht\gnBoxA%
    \ifnum     \gnArgHgt<\gnCornerHgt \gnArgHgt=0pt%
    \else \advance \gnArgHgt by -\gnCornerHgt%
    \fi \raise\gnArgHgt\hbox{$\ulcorner$} \box\gnBoxA %
    \raise\gnArgHgt\hbox{$\urcorner$}}
\newcommand{\code}{\mathsf{code}}
\newcommand{\Det}{\mathsf{Det}}
\newcommand{{\cell}}{{\mathsf{cell}}}
\newcommand{\oes}{\sim}

@@

We describe a simple type theory and operational semantics for Idealized Algol with countable nondeterminism.
This is similar to the approach adopted in \cite{mcCHFiniteND}, which extends Idealized Algol with \emph{finite} nondeterminism.  
The types of our language are defined inductively as follows:
\[
  T \Coloneqq \nat \mid \com \mid \Var \mid T \to T\,.
  \]
Meanwhile, the terms are those given in \cite{SamsonGuyIAPassive}, together with the nondeterministic choice:
\begin{IEEEeqnarray*}{RCL}
  M & \Coloneqq & x \mid \lambda x . M \mid M\;M \mid \Y_T \mid \\
  && \n \mid \skipp \mid \suc \mid \pred \mid \\
  && \IfO \mid \blank;\blank \mid \blank \coloneqq \blank \mid \\
  && \deref\footnote{That is, variable \deref{}ccess.} \mid \neww_T \mid \mkvar \mid \wn\,.
\end{IEEEeqnarray*}

@@

We will use the Hyland-Ong version of Game Semantics, as in \cite{SamsonGuyIAPassive}.

@@

Our base arenas will be the \emph{flat arenas} for the types $\nat$ and $\com$.  
Given a set $X$, the flat arena on $X$ is the arena with a single $O$-question $q$ and a $P$-answer $x$ for each $x\in X$, where $*\ts q$ and $q\ts x$ for each $x$.  
The denotation of the type $\nat$ will be the flat arena $\bN$ on the set of natural numbers, while the denotation of the type $\com$ will be the flat arena $\bC$ on the singleton set $\{a\}$.

We shall assume that our arenas are \emph{enumerated}; that is, that the set $M_A$ comes equipped with a partial surjection $\bN\to M_A$. 
We assume that this enumeration is defined on the flat arenas and extended to the connectives in a natural way.

Given an arena $A$, a \emph{justified string} in $A$ is a sequence $s$ of moves in $A$, together with \emph{justification pointers} that go from move to move in the sequence.  
The justification pointers must be set up in such a way that every non-initial move $m$ in $s$ has exactly one justification pointer going back to an earlier move $n$ in $s$ such that $n\ts_A m$.  
We say that $n$ \emph{justifies} $m$.  
It is easy to see that every justified string must begin with an initial move, an hence with an $O$-question.

@@

We follow the approach taken by Abramsky and McCusker \cite{SamsonGuyIAPassive} -- a middle road between the \emph{arenas} of Hyland and Ong and the \emph{games} of \cite{ajmPcf} that makes the linear structure more apparent.

@@

Our definition of a strategy will be modelled upon that given in \cite{mcCHFiniteND}.  
We model nondeterministic computations by relaxing the determinism constraint on strategies -- so player $P$ may have multiple replies to any given $O$-move.

In addition, we have to keep track of any possible divergence in the computation; this is so we can distinguish terms such as
\begin{mathpar}
  \IfO \wn\;\Omega\;\0 \and \0\,,
\end{mathpar}
where the term on the right must converge (to $\0$), while the term on the left has a possible divergence.  
The traditional way of representing divergences in game semantics is by a partiality in the strategy; i.e., an $O$-move to which $P$ has no reply, but this doesn't work here: $\IfO\wn\;\Omega\;\0$ and $\0$ both denote the strategy with maximal play $q\;0$; the partiality in the former is `hidden' by the move $0$.

To fix this problem, we follow \cite{mcCHFiniteND} by modelling a strategy as a pair $(T_\sigma,D_\sigma)$, where $T_\sigma$ is a nondeterministic strategy in the usual sense and $D_\sigma$ is a set of $O$-plays after which there is a possibility of divergence.

Tracking divergences explicitly in this way requires some care when we compose strategies using `parallel composition plus hiding'.
Specifically, we need to be able to add new divergences into strategies when they arise through `infinite chattering' or \emph{livelock}.  
For example, the denotation of the term
\[
  M = \Y_{\nat\to\nat}(\lambda f.\lambda n.n;(f n))
  \]
is given by a total strategy, without divergences: namely the strategy $\mu$ with plays of the form shown in Figure \ref{fig:mn-plays}(a).
\begin{figure}
  \begin{mathpar}
    a) \and
    \begin{array}{cc}
      \mathbf{\bN_1} & \mathbf{\bN_2} \\
                     & q \\
                  q  &   \\
                 n_1 &   \\
                  q  &   \\
                 n_2 &   \\
              \vdots &   \\
              {}     & {}
    \end{array}
    \and b) \and
    \begin{array}{cc}
      \mathbf{\bN_1} & \mathbf{\bN_2} \\
                     & q \\
                  q  &   \\
                 n_1 &   \\
              \vdots &   \\
                  q  &   \\
                 n_m &   \\
                     & 0 \\
    \end{array}
  \end{mathpar}
  \caption{Finite plays alone are not sufficient to distinguish between terms of a language with countable nondeterminism.}
  \label{fig:mn-plays}
\end{figure}
However, when we compose this strategy with any total strategy for $\bN$ on the left, we expect the resulting strategy to contain divergences, since the term $M \n$ diverges for any $\n$.
Semantically, this corresponds to the fact that we have a legal interaction $q\;q\;n\;q\;n\;\cdots$ with an infinite tail in $\bN_1$; when we perform `hiding' by restricting the interaction to $\bN$, we have no reply to the initial move $q$.

@@

Harmer and McCusker's approach works very satisfactorily for finite nondeterminism, but not at all for countable nondeterminism.  
To see why, consider the term
\[
  N = \Y_{\nat\to\nat\to\nat}(\lambda g.\lambda m n.\IfO m\;\0\;n;(g\;(\pred m)\;n))\wn\,.
  \]
This term first chooses a natural number $m$, and then reads from its input $n$ for a total of $m$ times before eventually returning $\0$.  
Thus, its denotation is the strategy $\nu$ with maximal plays of arbitrary length of the form shown in Figure \ref{fig:mn-plays}(b).
Note that this strategy strictly contains the strategy $\mu$ that we considered before, and therefore that the denotation of
\[
  \IfO \wn M N
  \]
has the same denotation as $N$, even though for any $n$, $M\n\not{\mustconverge}$, while $N\n\mustconverge$.
Moreover, if we try to compose $\deno{N}$ with the strategy on $\bN$ that always returns $1$, then we end up with an infinite increasing sequence of positions, which triggers the introduction of a divergent play into the composite strategy -- even though no divergence occurs in the evaluation of $N$.

Aside from showing that this model cannot possibly be sound, this example actually leads to composition not being associative if we naively extend the Harmer-McCusker model from finite to infinite nondeterminism (e.g., see \cite[4.4.1]{RusssThesis}).

Somehow, the crucial point is that we need to distinguish between terms like $M$, which contain infinite sequences of moves, and terms like $N$, which contain arbitrarily large finite sequences of moves.  
The way that we do this is by making the infinite sequences of moves explicit in our strategies, in the style of \cite{RoscoeCspInfinite} and \cite{LevyGsInfinite}.  
When we use this technique, the denotation of $M$ will contain an infinite sequence, while the denotation of $N$ will contain arbitrarily long finite sequences, but no infinite sequences.

The games in our model are the same as those that we considered in the last section, but our definition of a strategy will change.

@@

Given an arena $A$, we define an \emph{infinite justified string} in the obvious way.  
We say that such a string is \emph{recursive} if it corresponds, via the enumeration on $M_A$ to a recursive function $\bN\to\bN$.

We define $\overline{P_A}$ to be $P_A$ together with the set of all those recursive infinite justified sequences that have all finite prefixes in $P_A$.

We deliberately ignore any non-recursive infinitary behaviours, since these cannot be detected by computable contexts.

Let $A$ be a game.  
A \emph{strategy} $\sigma$ for $A$ is a pair $(T_\sigma,D_\sigma)$, where:
\begin{itemize}
  \item $T_\sigma$ is a non-empty prefix-closed subset of $\overline{P_A}$ such that if $s\in T_\sigma$ is a $P$-position and $sa\in P_A$ then $sa\in T_\sigma$.
  \item $D_\sigma\subset \overline{P_A}$ is a postfix-closed set of plays in $\overline{P_A}$ that either end with an $O$-move or are infinite.  
    We require $D_\sigma$ to obey the following rules:
    \begin{description}
      \item[Divergences come from plays] If $d\in D_\sigma$ then there exists some $s\prefix d$ such that $s\in T_\sigma\cap D_\sigma$.
      \item[Diverge-or-reply] If $s\in T_\sigma$ is an $O$-position, then either $s\in D_\sigma$ or $sa\in T_\sigma$ for some legal play $sa$.
      \item[Infinite positions are divergent] If $s\in T_\sigma$ is infinite, then $s\in D_\sigma$.
    \end{description}
\end{itemize}

@@

We call the sequence $\s$ a \emph{legal interaction} if $\s\vert_{A,B}\in P_{A\implies B}$, $\s\vert_{B,C}\in P_{B\implies C}$ and $\s\vert_{A,C}\in P_{A\implies C}$.  
We write $\Int_\infty(A,B,C)$ for the set of (possibly infinite) legal interactions between $A$, $B$ and $C$.

Now, given strategies $\sigma\from A \implies B$ and $\tau\from B \implies C$, we define
\[
  T_\sigma\|T_\tau = \{\s\in\Int_\infty(A,B,C)\suchthat \s\vert_{A,B}\in T_\sigma,\;\s\vert_{B,C}\in T_\tau\}\,,
  \]
and then set
\[
  T_{\sigma;\tau} = \{\s\vert_{A,C}\suchthat \s\in T_\sigma\|T_\tau\}\,.
  \]
As for divergences in $\sigma;\tau$, our approach is actually simpler than that in \cite{mcCHFiniteND}; we set
\[
  D_\sigma\dv D_\tau = \left\{\s\in \Int_\infty(A,B,C)\;\middle|\; \mbox{\pbox{\textwidth}{\textbf{either} $\s\vert_{A,B}\in D_\sigma$ and $s\vert_{B,C}\in T_\tau$ \\ \textbf{or} $\s\vert_{A,B}\in T_\sigma$ and $s\vert_{B,C}\in D_\tau$}}\right\}\,.
  \]
We then set
\[
  D_{\sigma;\tau} = \pocl_{A\implies C}\{\s\vert_{A,C}\suchthat\s\in D_\sigma\dv D_\tau\}\,,
  \]
where $\pocl X$ denotes the \emph{postfix closure} of $X$; i.e., the set of all $O$-plays in $P_{A\implies C}$ that have some prefix in $X$.

Note that there is no need to consider separately, as Harmer and McCusker do, divergences that arise through `infinite chattering': in our model, a case of infinite chattering between strategies $\sigma$ and $\tau$ is itself an (infinite) legal interaction between the two strategies, which is necessarily divergent (because it is infinite) and therefore gives rise to some divergence in $\sigma;\tau$.

@@

Any finite-nondeterminism strategy in the sense of \cite{mcCHFiniteND} may be interpreted as a complete strategy by enlarging it with all its infinite limiting plays.  
However, when we introduce countable nondeterminism, we also introduce strategies that are not complete.  
For example, the strategy $\nu$ that we mentioned above has an infinite increasing sequence of plays $q0\prefix q0q0\prefix \cdots$, but has no infinite play corresponding to its limit.  
Nonetheless, we do not want to allow arbitrary strategies: for example, the strategy $\mu$ above should include the infinite play $qq0q0\dots$; the strategy $\mu^\circ$ formed by removing this infinite play has no meaning in our language.  
Indeed, if we compose $\mu^\circ$ with the strategy $\0$ for $\bN$ on the left, then the resulting strategy does not satisfy diverge-or-reply.
The difference with $\nu$ is that every play $qq0\cdots q0\in T_\nu$ may be completed in $\nu$ by playing the move $0$ on the right.
In other words, $\nu$ is the union of complete strategies, while $\mu^\circ$ is not.

@@

\begin{definition}
  We say that a strategy $\sigma$ for a game $A$ is \emph{deterministic} if
  \begin{itemize}
    \item it is complete;
    \item whenever $sab,sac$ are $P$-plays in $T_\sigma$ we have $b=c$ and the justifier of $b$ is the justifier of $c$;
    \item If $s\in D_\sigma$ then either $s$ is infinite or there is no $a$ such that $sa\in T_\sigma$.
  \end{itemize}
\end{definition}

\begin{lemma}
  Let $A,B,C$ be games and let $\sigma\from A\implies B$, $\tau\from B\implies C$ be deterministic complete strategies.  
  Then $\sigma;\tau$ is complete.
  \label{lem:complete-deterministic-composition}
\end{lemma}
\begin{proof}
  The proof relies on a lemma from \cite{hoPcf} that states (in our language) that if $\sigma$ and $\tau$ are deterministic strategies and $s\in T_{\sigma;\tau}$ then there is a unique minimal $\s\in T_\sigma\|T_\tau$ such that $\s\vert_{A,C}=s$.  
  That means that if $s_1\prefix s_2\prefix \cdots$ is an infinite increasing sequence of plays in $T_{\sigma;\tau}$, with infinite limit $s$, then there is a corresponding infinite increasing sequence of legal interactions $\s_1 \prefix \s_2 \prefix \cdots$.  
  Then the limit of this sequence is an infinite legal interaction $\s$ and we must have $\s\vert_{A,B}\in\sigma$, $\s\vert_{B,C}\in\tau$ by completeness of $\sigma$ and $\tau$.  
  Therefore, $s=\s\vert_{A,C}\in T_{\sigma;\tau}$.
\end{proof}

\begin{corollary}
  The composition of strategies $\sigma\from A\implies B$ and $\tau\from B \implies C$ is a well-formed strategy for $A\implies C$.
  \label{cor:diverge-or-reply}
\end{corollary}
\begin{proof}
  The only tricky point is establishing that diverge-or-reply holds for $\sigma;\tau$.  
  Again, it is sufficient to prove this in the case that $\sigma$ and $\tau$ are deterministic and complete.
  Then it essentially follows from the argument used in \cite{abramskyjagadeesangames} that shows that a partiality at an $O$-position $s\in T_{\sigma;\tau}$ must arise either from a partiality in $T_\sigma$ or $T_\tau$ or from `infinite chattering' between $\sigma$ and $\tau$.  
  In the first case, the diverge-or-reply rule for $\sigma$ and $\tau$ gives us a divergence at $s$ in $\sigma;\tau$.  
  In the second case, an infinite chattering between $\sigma$ and $\tau$ corresponds to an infinite interaction $\s\in\Int(A,B,C)$ ending with infinitely many moves in $B$ such that $\s\vert_{A,C}=s$. 
  Completeness for $\sigma$ and $\tau$ tells us that $\s\vert_{A,B}\in D_\sigma$ and $\s\vert_{B,C}\in D_\tau$ and therefore that $\s\vert_{A,C}\in D_{\sigma;\tau}$.  
\end{proof}

>>>>

\let\oldemptyset\emptyset
\let\emptyset\varnothing
\newcommand{\nat}{{\mathtt{nat}}}
\newcommand*\from{\colon}
\newcommand{\Y}{\mathbf{Y}}
\newcommand{\opto}{\longrightarrow}
\newcommand{\n}{{\mathtt{n}}}
\newcommand{\IfO}{{\mathtt{If0}\;}}
\newcommand{\suc}{{\mathtt{suc\;}}}
\newcommand{\pred}{{\mathtt{pred\;}}}
\newcommand{\0}{{\mathtt{0}}}
\newcommand{\com}{{\mathtt{com}}}
\newcommand{\skipp}{{\mathsf{skip}}}
\newcommand{\neww}{{\mathsf{new}}}
\newcommand{\mkvar}{{\mathsf{mkvar}\;}}
\newcommand{\deref}{\texttt{@}}
\newcommand{\ia}[2]{\langle #1 , #2 \rangle}
\newcommand{\stup}[3]{\langle #1 \mid #2 \mapsto #3 \rangle}
\newcommand{\Var}{{\mathtt{Var}}}
\newcommand{\new}{{\mathtt{new}}}
\newcommand{\case}{{\mathtt{case}}}
\newcommand{\blank}{\;\underline{\hspace{1.5ex}}\;}
\newcommand{\bN}{\mathbb{N}}
\newcommand{\bC}{\mathbb{C}}
\newcommand{\deno}[1]{\left\llbracket#1\right\rrbracket}
\newcommand{\converges}{\Downarrow}
\newcommand{\mustconverge}{\converges^{\text{must}}}
\newcommand{\Iflt}{\mathtt{If{<}\;}}
\newcommand{\Ifgt}{\mathtt{If{>}\;}}
\newcommand{\inr}{{\mathsf{inr}}}
\newcommand{\inl}{{\mathsf{inl}}}
\newcommand{\lmam}{\sqsubseteq_{m\&m}}
\newcommand{\emam}{\equiv_{m\&m}}
\newcommand{\lst}{\lesssim}
\newcommand{\rulename}[1]{\LeftTirNameStyle{#1}}
\newcommand{\ts}{{\;\vdash}}
\newcommand{\nts}{{\;\not\vdash}}
\newcommand{\OP}{\{O,P\}}
\newcommand{\QA}{\{Q,A\}}
\DeclareMathOperator{\id}{id}
\DeclareMathOperator{\pr}{pr}
\newcommand{\suchthat}{\;\colon\;}
\newcommand{\varsuchthat}{\;\mid\;}
\newcommand{\esuchthat}{\;.\;}
\newcommand{\emptyplay}{\epsilon}
\newcommand{\tensor}{\otimes}
\renewcommand{\implies}{\multimap}
\newcommand{\prefix}{\sqsubseteq}
\newcommand{\s}{\mathfrak s}
\newcommand{\C}{{\mathcal{C}}}
\newcommand{\G}{\mathcal G}
\DeclareMathOperator{\Int}{int}
\newcommand{\dv}{{\lightning}}
\DeclareMathOperator{\pocl}{pocl}
\newbox\gnBoxA
\newdimen\gnCornerHgt
\setbox\gnBoxA=\hbox{$\ulcorner$}
\global\gnCornerHgt=\ht\gnBoxA
\newdimen\gnArgHgt
\def\pv #1{%
    \setbox\gnBoxA=\hbox{$#1$}%
    \gnArgHgt=\ht\gnBoxA%
    \ifnum     \gnArgHgt<\gnCornerHgt \gnArgHgt=0pt%
    \else \advance \gnArgHgt by -\gnCornerHgt%
    \fi \raise\gnArgHgt\hbox{$\ulcorner$} \box\gnBoxA %
    \raise\gnArgHgt\hbox{$\urcorner$}}
\newcommand{\code}{\mathsf{code}}
\newcommand{\Det}{\mathsf{Det}}
\newcommand{{\cell}}{{\mathsf{cell}}}
\newcommand{\oes}{\sim}

@@

We describe a type theory and operational semantics for Idealized Algol with countable nondeterminism.
The types of our language are defined inductively as follows:
\[
  T \Coloneqq \nat \mid \com \mid \Var \mid T \to T\,.
  \]
Meanwhile, the terms are those given in \cite{SamsonGuyIAPassive}, together with the nondeterministic choice:
\begin{IEEEeqnarray*}{RCL}
  M & \Coloneqq & x \mid \lambda x . M \mid M\;M \mid \Y_T \mid \\
  && \n \mid \skipp \mid \suc \mid \pred \mid \\
  && \IfO \mid \blank;\blank \mid \blank \coloneqq \blank \mid \\
  && \deref\footnote{That is, variable \deref{}ccess.} \mid \neww_T \mid \mkvar \mid \wn\,.
\end{IEEEeqnarray*}

@@

@@

Our base arenas will be the \emph{flat arenas} for the types $\nat$ and $\com$.  
Given a set $X$, the flat arena on $X$ is the arena with a single $O$-question $q$ and a $P$-answer $x$ for each $x\in X$, where $*\ts q$ and $q\ts x$ for each $x$.  
The denotations of the types $\nat$ and $\com$ are the flat arenas $\bN$ and $\bC$ on, respectively, the set of natural numbers and the singleton $\{a\}$.

We assume that our arenas are \emph{enumerated}; i.e., that the set $M_A$ is equipped with a partial surjection $\bN\to M_A$.  
The denotation of any IA type has a natural enumeration.

Given an arena $A$, a \emph{justified string} in $A$ is a sequence $s$ of moves in $A$, together with \emph{justification pointers} that go from move to move in the sequence.  
The justification pointers must be set up in such a way that every non-initial move $m$ in $s$ has exactly one justification pointer going back to an earlier move $n$ in $s$ such that $n\ts_A m$; we say that $n$ \emph{justifies} $m$.  
In particular, every justified string begins with an initial move, and hence with an $O$-question.

@@

We use the approach taken by Abramsky and McCusker \cite{SamsonGuyIAPassive} -- a middle road between the \emph{arenas} of Hyland and Ong and the \emph{games} of \cite{ajmPcf} that makes the linear structure more apparent.

@@

As in \cite{mcCHFiniteND}, we model nondeterministic computations by relaxing the determinism constraint on strategies -- so player $P$ may have multiple replies to any given $O$-move.

In addition, we have to keep track of any possible divergence in the computation so we can distinguish terms such as
$
  \IfO \wn\;\Omega\;\0$, which may diverge, and $\0$, which must converge.

To fix this problem, we follow \cite{mcCHFiniteND} by modelling a strategy as a pair $(T_\sigma,D_\sigma)$, where $T_\sigma$ is a nondeterministic strategy in the usual sense and $D_\sigma$ is the set of those $O$-positions where there is a possibility of divergence.

Tracking divergences explicitly in this way requires some care when we compose strategies.
Specifically, we need to be able to add new divergences into strategies when they arise through `infinite chattering' or \emph{livelock}.  
For example, the denotation of the term
\[
  M = \Y_{\nat\to\nat}(\lambda f.\lambda n.n;(f n))
  \]
is given by a total strategy, without divergences: namely the strategy $\mu$ with plays of the form shown in Figure \ref{fig:mn-plays}(a).
\begin{figure}
  \begin{mathpar}
    a) \and
    \begin{array}{cc}
      \mathbf{\bN_1} & \mathbf{\bN_2} \\
                     & q \\
                  q  &   \\
                 n_1 &   \\
                  q  &   \\
                 n_2 &   \\
              \vdots &   \\
              {}     & {}
    \end{array}
    \and b) \and
    \begin{array}{cc}
      \mathbf{\bN_1} & \mathbf{\bN_2} \\
                     & q \\
                  q  &   \\
                 n_1 &   \\
              \vdots &   \\
                  q  &   \\
                 n_m &   \\
                     & 0 \\
    \end{array}
  \end{mathpar}
  \caption{Finite plays alone are not sufficient to distinguish between terms of a language with countable nondeterminism.}
  \label{fig:mn-plays}
\end{figure}
However, when we compose this strategy with any total strategy for $\bN$ on the left, we expect the resulting strategy to contain divergences, since the term $M \n$ diverges for any $\n$.
Semantically, this corresponds to the fact that we have a legal interaction $q\;q\;n\;q\;n\;\cdots$ with an infinite tail in $\bN_1$; when we perform `hiding' by restricting the interaction to $\bN$, we have no reply to the initial move $q$.

@@

Harmer and McCusker's approach works very satisfactorily for finite nondeterminism, but not at all for countable nondeterminism.  
To see why, consider the term
\[
  N = \Y_{\nat\to\nat\to\nat}(\lambda g.\lambda m n.\IfO m\;\0\;n;(g\;(\pred m)\;n))\wn\,.
  \]
This term first chooses a natural number $m$, and then reads from its input $n$ for a total of $m$ times before eventually returning $\0$.  
Thus, its denotation is the strategy $\nu$ with maximal plays of arbitrary length of the form shown in Figure \ref{fig:mn-plays}(b).
Note that this strategy strictly contains the strategy $\mu$ that we considered before, and therefore that the denotation of
  $\IfO \wn M N$
has the same denotation as $N$, even though for any $n$, $M\n\not\mustconverge$, while $N\n\mustconverge$.
Moreover, if we try to compose $\deno{N}$ with the strategy on $\bN$ that always returns $1$, then we end up with an infinite increasing sequence of positions, which triggers the introduction of a divergent play into the composite strategy -- even though $N$ must converge.

Aside from showing that the naive extension of the Harmer-McCusker model cannot be sound, this example actually leads to composition not being associative (e.g., see \cite[4.4.1]{RusssThesis}).

What this illustrates is that we can no longer deduce the infinitary behaviour of a strategy by looking at the limits of its finite plays; instead, we need to keep track of infinite sequences of moves explicitly, in the style of \cite{RoscoeCspInfinite} and \cite{LevyGsInfinite}.
When we use this technique, the denotation of $M$ will contain an infinite sequence, while the denotation of $N$ will contain arbitrarily long finite sequences, but no infinite sequences.

@@

We define an \emph{infinite justified string} in an arena $A$ in the obvious way.  
We say such a string is \emph{recursive} if it corresponds, via the enumeration on $M_A$, to a recursive function $\bN\to\bN$.

We define $\overline{P_A}$ to be $P_A$ together with the set of all those recursive infinite justified sequences that have all finite prefixes in $P_A$.
Note that we deliberately ignore any non-recursive infinitary behaviours, since these cannot be detected by computable contexts.

Let $A$ be a game.  
A \emph{strategy} $\sigma$ for $A$ is a pair $(T_\sigma,D_\sigma)$, where:
\begin{itemize}
  \item $T_\sigma$ is a non-empty prefix-closed subset of $\overline{P_A}$ such that if $s\in T_\sigma$ is a $P$-position and $sa\in P_A$ then $sa\in T_\sigma$.
  \item $D_\sigma\subset \overline{P_A}$ is a postfix-closed set of plays in $\overline{P_A}$ that either end with an $O$-move or are infinite.  
    We require $D_\sigma$ to obey the following rules:
    \begin{description}
      \item[Divergences come from plays] If $d\in D_\sigma$ then there exists $s\prefix d$ such that $s\in T_\sigma\cap D_\sigma$.
      \item[Diverge-or-reply] If $s\in T_\sigma$ is an $O$-position, then either $s\in D_\sigma$ or $sa\in T_\sigma$ for some $sa$.
      \item[Infinite positions are divergent] If $s\in T_\sigma$ is infinite, then $s\in D_\sigma$.
    \end{description}
\end{itemize}

@@

We call $\s$ a \emph{legal interaction} if $\s\vert_{A,B}\in P_{A\implies B}$, $\s\vert_{B,C}\in P_{B\implies C}$ and $\s\vert_{A,C}\in P_{A\implies C}$.  
We write $\Int_\infty(A,B,C)$ for the set of (possibly infinite) legal interactions between $A$, $B$ and $C$.

Now, given strategies $\sigma\from A \implies B$ and $\tau\from B \implies C$, we define
\[
  T_\sigma\|T_\tau = \{\s\in\Int_\infty(A,B,C)\suchthat \s\vert_{A,B}\in T_\sigma,\;\s\vert_{B,C}\in T_\tau\}\,,
  \]
and then set
  $T_{\sigma;\tau} = \{\s\vert_{A,C}\suchthat \s\in T_\sigma\|T_\tau\}$.

As for divergences in $\sigma;\tau$, our approach is actually simpler than that in \cite{mcCHFiniteND}; we set
\[
  D_\sigma\dv D_\tau = \left\{\s\in \Int_\infty(A,B,C)\;\middle|\; \mbox{\pbox{\textwidth}{\textbf{either} $\s\vert_{A,B}\in D_\sigma$ and $s\vert_{B,C}\in T_\tau$ \\ \textbf{or} $\s\vert_{A,B}\in T_\sigma$ and $s\vert_{B,C}\in D_\tau$}}\right\}\,.
  \]
We then set
  $D_{\sigma;\tau} = \pocl_{A\implies C}\{\s\vert_{A,C}\suchthat\s\in D_\sigma\dv D_\tau\}$,
where $\pocl X$ denotes the \emph{postfix closure} of $X$; i.e., the set of all $O$-plays in $P_{A\implies C}$ that have some prefix in $X$.

Note that there is no need to consider separately, as Harmer and McCusker do, divergences that arise through `infinite chattering': in our model, a case of infinite chattering between strategies $\sigma$ and $\tau$ is itself a legal interaction between the two strategies, which is necessarily divergent (because it is infinite) and therefore gives rise to some divergence in $\sigma;\tau$.

@@

Any finite-nondeterminism strategy in the sense of \cite{mcCHFiniteND} may be interpreted as a complete strategy by enlarging it with all its infinite recursive limiting plays.  
However, when we introduce countable nondeterminism, we also introduce strategies that are not complete.  
For example, the strategy $\nu$ that we mentioned above has an infinite increasing sequence of plays $q0\prefix q0q0\prefix \cdots$, but has no infinite play corresponding to its limit.  
Nonetheless, we do not want to allow arbitrary strategies: for example, the strategy $\mu$ above should include the infinite play $qq0q0\dots$; the strategy $\mu^\circ$ formed by removing this infinite play has no meaning in our language.  
Indeed, if we compose $\mu^\circ$ with the strategy $\0$ for $\bN$ on the left, then the resulting strategy does not satisfy diverge-or-reply.
The difference with $\nu$ is that every play $qq0\cdots q0\in T_\nu$ may be completed in $\nu$ by playing the move $0$ on the right.
In other words, $\nu$ is the union of complete strategies, while $\mu^\circ$ is not.

@@

\begin{definition}
  We say that a strategy $\sigma$ for a game $A$ is \emph{deterministic} if
  \begin{itemize}
    \item it is complete;
    \item if $sab,sac$ are $P$-plays in $T_\sigma$ then $b=c$ and the justifier of $b$ is the justifier of $c$;
    \item if $s\in D_\sigma$ then either $s$ is infinite or there is no $a$ such that $sa\in T_\sigma$.
  \end{itemize}
\end{definition}

\begin{lemma}
  Let $A,B,C$ be games and let $\sigma\from A\implies B$, $\tau\from B\implies C$ be deterministic strategies.  
  Then $\sigma;\tau$ is complete.
  \label{lem:complete-deterministic-composition}
\end{lemma}
\begin{proof}
  The proof relies on a lemma from \cite{hoPcf} that states (in our language) that if $\sigma$ and $\tau$ are deterministic strategies and $s\in T_{\sigma;\tau}$ then there is a unique minimal $\s\in T_\sigma\|T_\tau$ such that $\s\vert_{A,C}=s$.  
  That means that if $s_1\prefix s_2\prefix \cdots$ is an infinite increasing sequence of plays in $T_{\sigma;\tau}$, with limit $s$, then there is a corresponding infinite increasing sequence of legal interactions $\s_1 \prefix \s_2 \prefix \cdots$.  
  Then the limit of the $\s_i$ is an infinite legal interaction $\s$ and we must have $\s\vert_{A,B}\in\sigma$, $\s\vert_{B,C}\in\tau$ by completeness of $\sigma$ and $\tau$.  
  Therefore, $s=\s\vert_{A,C}\in T_{\sigma;\tau}$.
\end{proof}

\begin{corollary}
  The composition of strategies $\sigma\from A\implies B$ and $\tau\from B \implies C$ is a well-formed strategy for $A\implies C$.
  \label{cor:diverge-or-reply}
\end{corollary}
\begin{proof}
  The only tricky point is establishing that diverge-or-reply holds for $\sigma;\tau$.  
  Again, it is sufficient to prove this in the case that $\sigma$ and $\tau$ are deterministic and complete.
  Then it essentially follows from the argument used in \cite{abramskyjagadeesangames} that shows that a partiality at an $O$-position $s\in T_{\sigma;\tau}$ must arise either from a partiality in $T_\sigma$ or $T_\tau$ or from `infinite chattering' between $\sigma$ and $\tau$.  
  In the first case, the diverge-or-reply rule for $\sigma$ and $\tau$ gives us a divergence at $s$ in $\sigma;\tau$.  
  In the second case, an infinite chattering between $\sigma$ and $\tau$ corresponds to an infinite interaction $\s\in\Int_\infty(A,B,C)$ (with a tail in $B$) such that $\s\vert_{A,C}=s$.  Completeness for $\sigma$ and $\tau$ tells us that $\s\vert_{A,B}\in D_\sigma$ and $\s\vert_{B,C}\in D_\tau$ and therefore that $\s\vert_{A,C}\in D_{\sigma;\tau}$.  \end{proof}

Our proof for Corollary \ref{cor:diverge-or-reply} really makes use of the fact that a locally complete strategy is \emph{lively} in the sense of \cite{LevyGsInfinite}; i.e., locally deterministic.  
Our definition is slightly stronger than liveliness, because it insists that the union of complete strategies be countable.
This will be essential to our definability result.
